---
title: "gapminder"
author: "Leslie"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
gapminder <- read_csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
```

```{r}
head(gapminder, 10)
```

```{r}
str(gapminder)
```

```{r}
dim(gapminder)
ncol(gapminder)
nrow(gapminder)
```

```{r}
unique(gapminder[,"country"])
```

```{r}
summary(gapminder)
```


```{r}
# install.packages("skimr")
```

```{r}
library(skimr)
```

```{r}
skim(gapminder)
```

```{r}
dplyr::filter(gapminder, country == "Switzerland")
```

```{r}
dplyr::filter(gapminder, country == "Afghanistan")
```

```{r}
dplyr::select(gapminder, country, year, pop)
```

```{r}
gapminder_no_cont <- dplyr::select(gapminder, -continent)
dplyr::filter(gapminder_no_cont, year == 2007)
```

##Pipe.
Applies part after pipe to the selection before the pipe

```{r}
select(gapminder, -continent) %>% 
  filter(year == 2007)
```

##Mutate
uses data in one column to make calculations and add a new column with the results. Ex: "gdp" is new column calculated from gdpPercap multiplied by population. Also can sort this using arrange()

```{r}
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  arrange(desc(gdp))
```

##Groupby

to use a subset of the data. good to ungroup at the end

```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = gdpPercap * pop, max_gdp = mad(gdp)) %>%
  ungroup()
```

#Combining groub_by and summarize

```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarise(max_gdp = max(gdp))
```

#Joining Datasets
how to combine two datasets  
multiple ways, left_join, right_join, inner_join, full_join   

Adding a second dataset of co2 observations (just a subset)
```{r}
co2 <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```
```{r}
head(co2)
summary(co2)
```

joining c02 data to gapminder data   
first select only 2007 data from gapminder
```{r}
gap_2007 <- gapminder %>% 
  filter(year == 2007) 
```

#left_join

```{r}
left_join(gap_2007, co2, by = "country")
```

#right_join
```{r}
right_join(gap_2007, co2, by = "country")
```

##Tidyr


```{r}
gap_wide <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")
```

inserting - means you wont use that column (exclude)

```{r}
gap_wide %>%
  gather(observation, observ_value, -continent, -country)
```

# to separate the year from the observation to tidy up the dataset  
also can convert the year column from a character to an integer

```{r}
gap_wide %>%
  gather(observation, observ_value, -continent, -country) %>%
  separate(observation, into = c("obs_type", "year"), sep = "_", convert = TRUE)
```
create a new variable of this dataset
```{r}
gap_better <- gap_wide %>%
  gather(observation, observ_value, -continent, -country) %>%
  separate(observation, into = c("obs_type", "year"), sep = "_", convert = TRUE)
```



```{r}
unique(gap_better$obs_type)
```

```{r}
gap_better %>%
  spread(obs_type, observ_value)
```






